Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-07 numerical.tex
6.2 edited.

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. % Note that for the fiducial disk model the critical thermal 
% timescale $\beta_\mathrm{crit} = 0.125$.
Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta$ from $0.01$ to  $0.1$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.  %analytic
% results for low order modes

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as higher order modes have larger
growth rates. As noted above, these higher order modes are affected by
$\zmax$, e.g. there is a significant mismatch between numerical and 
analytical results for $M\geq1$ when $\zmax=3H$ (upper panel).

%However, by comparing with results for $\zmax=3H$, we see that

However, the fundamental mode becomes dominant for $\beta \geq 
0.03$. Higher order modes, which would be dominant at small $\beta$, 
are more effectively stabilized with increasing $\beta$. For 
$\beta=0.1$, which is close to the critical thermal timescale
$\beta_\mathrm{crit} (=0.125$ for the fiducial disk
parameters), we only find the fundamental mode. This is consistent with
the discussion in \S\ref{iso_vsi_beta_crit}.

@@

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. Comparing these plots with
Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}, which have
$\beta=10^{-3}$, we see that increasing $\beta$ restricts the region 
characterized by $W\sim z$ and $\delta v_z\sim$ constant closer to the
midplane, with oscillatory behavior emerging near vertical boundaries
where buoyancy first becomes important. This leads to the `tilted'
pressure perturbations seen in Fig. \ref{lowfreq_eigenfunc_2d_cool} 
(cf. Fig. \ref{lowfreq_eigenfunc_2d}).

@@

%presence of buoyancy forces at the boundary may make BC more
%influential

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable 
modes with $\khat=30$. For clarity, we only plot analytic results 
for the fundamental mode ($M=0$) and $M=1,\,2$, and omit the 
squares. The fundamental mode is again insensitive to $\zmax$ and are
in close agreement with analytic results. However, 
there is increasing mismatch between analytical and numerical results
for higher order modes as $\beta$ is increased. This suggests that, in
the presence of buoyancy, boundary effects become important for these
modes, which have larger $\khat$ than that considered previously. For
example, for $\zmax=5H$ and $\zmax=7H$, we 
find modes slightly more unstable than the fundamental
mode even when $\beta=0.1$ (the leftmost eigenvalues in the lower two
panels), but these are clearly affected by boundary
conditions as they are absent for $\zmax=3H$.

Nevertheless, as before we observe more effective stabilization for
higher order modes, including the absence of surface modes, for
example as $\beta$ is increased from $0.01$ to $0.03$ (for $\zmax\geq
5H$). 
%\emph{not a clear and possible not valid argument.  All modes are associated with
%a BC and the disappearance of a mode for smaller zmax could be the artificial effect.}
%If we discard modes associated with boundary conditions, then
%the fundamental mode would again become the dominant mode as $\beta$ is
%increased.

@@

% Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
% show that the fundamental mode is not sensitive to boundary
% conditions. The fundamental mode also becomes dominant as $\beta$
% approaches $\beta_\mathrm{crit}$, except with the possible emergence
% of modes associated with boundary conditions. Thus, tracking the
% fundamental mode is a more preferable assessment of stability as $\beta$
% is increased from zero, rather than searching for the most unstable
% mode. This is because the latter may correspond to modes that depend on boundary
% conditions, e.g. surface modes whose growth rate varies with
% $\zmax$ in a vertically isothermal disk.

%boundary conditions play an increasing role for higher M

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
In Fig. \ref{bcrit_compare1} we plot the maximum VSI growth rates
found in the fiducial disk model as a function of the thermal
relaxation timescale $\beta$ for several values of $\khat$ and
$\zmax$. We also plot the critical timescale
$\beta_\mathrm{crit}=0.125$ for the fiducial disk parameters.

>>>>

@@

\subsection{Slower thermal relaxation}\label{therm_relax_eff}
We now consider the effect of longer cooling timescales by gradually increasing 
$\beta$. We expect VSI growth for  $\beta < \beta_\mathrm{crit} = 0.125$.  We also expect that higher order modes will damp at even lower $\beta$ values, as argued in \S\ref{iso_vsi_beta_crit}.

Fig. \ref{compare_modes_cool_kx10} largely confirms these expectations by plotting eigenvalues for $\khat=10$ 
and $\beta$ from $0.01$ to  $0.1$. 
The analytic results from Eq. \ref{relax_disp} are now plotted as different curves for each mode order $M$, with $\beta$ varying along each curve.  
We see the standard increase in frequency, $|\omega|$, with mode order.

As expected, the higher order modes are preferentially damped.  For $\beta \geq 0.03$ the fundamental ($M = 0$) mode is the fastest growing. For $\beta = 0.1$, only the $M = 0$ mode grows. This growth is slow since $\beta$ is near $\beta_\mathrm{crit}$.

Fig. \ref{compare_modes_cool_kx10} also shows how cooling times affect the dependence on $\zmax$, the size of the vertical domain.   For all cooling times, the fundamental mode depends only weakly on $\zmax$, and there is good agreement with analytic values.  This convergence is reassuring given the importance of the fundamental mode at longer cooling times.  For higher order modes and short cooling times, however, the eigenvalues vary strongly with $\zmax$.  The disagreement with analytic theory is strongest for the smallest domain, with $\zmax = 3 H$.  For $\beta \geq 0.03$, the $\zmax/H = 5, 7$ and the analytic results are well converged (aided by the fact that only $M < 5$ modes grow).

%We find the number of unstable modes decrease with increasing
%$\beta$. Consider the fiducial setup with $\zmax=5H$ 
%(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
%$\beta=0.01$ as higher order modes have larger
%growth rates. As noted above, these higher order modes are affected by
%$\zmax$, e.g. there is a significant mismatch between numerical and 
%analytical results for $M\geq1$ when $\zmax=3H$ (upper panel). 
%
%However, the fundamental mode becomes dominant for $\beta \geq 
%0.03$. Higher order modes, which would be dominant at small $\beta$, 
%are more effectively stabilized with increasing $\beta$. For 
%$\beta=0.1$, which is close to the critical thermal timescale
%$\beta_\mathrm{crit} (=0.125$ for the fiducial disk
%parameters), we only find the fundamental mode. This is consistent with
%the discussion in \S\ref{iso_vsi_beta_crit}.

@@

Figures \ref{lowfreq_eigenfunc_cool} and \ref{lowfreq_eigenfunc_2d_cool}
illustrate the eigenfunction of the fundamental mode with $\khat=10$ and
$\beta=0.1$. Compared with the small $\beta$ case in 
Figs. \ref{lowfreq_eigenfunc} and \ref{lowfreq_eigenfunc_2d}, the eigenfuctions 
show a more complex variation of phase with height.  This dependence yields 
the `tilted' appearance of pressure field in  Fig. \ref{lowfreq_eigenfunc_2d_cool}.
Physically, the increased role of buoyancy, which increases with height, explains 
why larger $\beta$ values produce more complex vertical structure.

@@

Fig. \ref{compare_modes_cool_kx30} shows how cooling affects higher wavenumber
modes, specifically $\khat=30$.   We focus on the $\zmax/H = 5, 7$ and analytic results for $\beta \geq 0.03$. 
These cases  show good convergence, though not quite as good as for $\khat = 10$.

With slower cooling, the wave frequency shows a more complex 
dependence on mode order, both analytically and numerically. For $\beta = 0.1$,
 the fundamental mode (which lies on the $M = 0$ curve) no longer has the smallest $|\omega|$ value.

Moreover, the fundamental mode is no longer the fastest growing mode for $\beta \geq 0.03$, or even 
for $\beta = 0.1$.  This complication is not actually surprising since $|hq\khat| = 1.5$ is no longer less than unity,
as required in the analytic derivation of \S\ref{iso_vsi_beta_crit}.  Though  inconvenient,  given that the 
fundemental mode is easy to identify and the most numerically converged, this complication is not ultimately significant
for the operability of the VSI.

Fig. \ref{compare_modes_cool_kx30} also demonstrates that artificial surface modes are damped for $\beta \geq 0.03$.

%For clarity, we only plot analytic results 
%for the fundamental mode ($M=0$) and $M=1,\,2$, and omit the 
%squares. The fundamental mode is again insensitive to $\zmax$ and are
%in close agreement with analytic results. However, 
%there is increasing mismatch between analytical and numerical results
%for higher order modes as $\beta$ is increased. This suggests that, in
%the presence of buoyancy, boundary effects become important for these
%modes, which have larger $\khat$ than that considered previously. For
%example, for $\zmax=5H$ and $\zmax=7H$, we 
%find modes slightly more unstable than the fundamental
%mode even when $\beta=0.1$ (the leftmost eigenvalues in the lower two
%panels), but these are clearly affected by boundary
%conditions as they are absent for $\zmax=3H$. 
%
%Nevertheless, as before we observe more effective stabilization for
%higher order modes, including the absence of surface modes, for
%example as $\beta$ is increased from $0.01$ to $0.03$ (for $\zmax\geq
%5H$).

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
Having explored the behavior of VSI modes with cooling, we turn to the numerical vailidity of the 
analytic cooling criterion, $\beta < \beta_\mathrm{crit} = h|q|/(\gamma -1)$.

...

In Fig. \ref{bcrit_compare1} we plot the maximum VSI growth rates
found in the fiducial disk model as a function of the thermal
relaxation timescale $\beta$ for several values of $\khat$ and
$\zmax$. We also plot the critical timescale
$\beta_\mathrm{crit}=0.125$ for the fiducial disk parameters.
